A Study of Common Bug Fix Patterns in Rust

### 2022-11-24 conclusion.tex
added changes

In this project, we developed Ruxanne, a tool for mining bug fix patterns in Rust. Using this tool, we mined the top 18 most starred Rust projects in Github to discover common bug fix patterns lied within their code changes. At the heart of Ruxanne, we used a novel code embedding approach to embed most important aspects of a code change in a fixed sized datapoint. We stored 87,726 datapoints drawn from 57,214 commits across these 18 projects. After using the DBSCAN clustering algorithm, and a subsequent manual analysis, we obtained 12 clusters of general fix patterns and 8 clusters of borrow-checker related fix patterns.

>>>>

In this project, we developed Ruxanne, a tool for mining bug fix patterns in Rust. Using this tool, we mined the top 18 most starred Rust projects in Github to discover common bug fix patterns within their code changes. At the heart of Ruxanne, we used a novel code embedding approach to embed the most important aspects of a code change in a fixed sized datapoint. We stored 87,726 datapoints drawn from 57,214 commits across these 18 projects. After using the DBSCAN clustering algorithm, and a subsequent manual analysis, we obtained 12 clusters of general fix patterns and 8 clusters of borrow-checker related fix patterns.

### 2022-11-24 discussion.tex
added changes

In this work, through Ruxanne we successfully mined 20 cross project bug fix patterns. We introduced these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect four of the BC-related patterns. To the best of our knowledge, the rest of the patterns cannot be reported by the current linting tools. That is why, we believe our work introduces a new direction for IDE tools to help Rust developers.

In each group, we categorized the bug fix patterns based on the underlying program element. The patterns encompass a wide variety of program elements both in the general and BC-related patterns. This shows that our weighting scheme is acceptably unbiased. Also, the introduced patterns are all cross-project fix patterns. Each fix pattern has been seen at least in three different projects, indicating its commonality.

Furthermore, we expected to obtain certain patterns regarding some certain Rust related features, such as \verb+move+, but our results did not include those patterns. Our speculation is that, since Rust has a strong compiler check before creating the program executable, the programmers usually get many errors, and they only push their code once it is free of compile errors. That is why, we think there are some patterns that were hidden from our sight.

@@

In our work there are three main threats to internal validity: (1) We use a weighting scheme which is based on a heuristic to give more values to tree elements closer to the leaves. In addition to that, we also manually adjusted the weights of borrow checker related elements to respect ownership framework's role in Rust, though we recorded their results separate from general results. As this approach is based on an intuition, it does not guarantee achieving optimal clusters. (2) The collected commits are the ones that fix a bug in a previously pushed commit. We do not consider the bugs that occur during a commit and also fixed during the same commit. (3) Our code embedding approach is based on the frequency of observed program elements in ASTs, and we use DBSCAN as our clustering algorithm. Other program code embedding methods and clustering algorithms (e.g. SLINK) might output new clusters that our pipeline is unable to find.

The threat to external validity can be the developers reporting the commits as bug fixing commits while they are not really fixing a functionality. Similarly, a commit message might not contain our target keywords while the commit is associated with bug fixing changes.

>>>>

In this work, using Ruxanne we successfully mined 20 cross project bug fix patterns. We introduced these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect four of the BC-related patterns. To the best of our knowledge, the rest of the patterns cannot be reported by the current linting tools. We therefore believe that our work introduces a new direction for IDE tools to help Rust developers.

In each group, we categorized the bug fix patterns based on the underlying program element. The patterns encompass a wide variety of program elements both in the general and BC-related patterns. This shows that our weighting scheme is not biased towards specific program elements. Also, our patterns are all cross-project fix patterns: each fix pattern has been seen at least in three different projects.

We expected to obtain certain patterns regarding some certain Rust related features, such as \verb+move+, but our results did not include those patterns. Our speculation is that, since Rust has a strong compiler check before creating the program executable, the programmers usually get many errors, and they only push their code once it is free of compile errors. That is why, we think there are some patterns that were hidden from our sight.

@@

In our work there are three main threats to internal validity: (1) We use a weighting scheme which is based on a heuristic to give more values to tree elements closer to the leaves. In addition to that, we also manually adjusted the weights of borrow checker related elements to respect the ownership framework's role in Rust, though we recorded their results separate from general results. As this approach is based on intuition, it does not guarantee achieving optimal clusters. (2) The collected commits are the ones that fix a bug in a previously pushed commit. We do not consider the bugs that occur during a commit and also fixed during the same commit. (3) Our code embedding approach is based on the frequency of observed program elements in ASTs, and we use DBSCAN as our clustering algorithm. Other program code embedding methods and clustering algorithms (e.g. SLINK) might output new clusters that our pipeline is unable to find.

A threat to external validity can be the developers reporting the commits as bug fixing commits while they are not really fixing a functionality. Similarly, a commit message might not contain our target keywords while the commit is associated with bug fixing changes.
\todo[inline]{Are internal and external used properly?}

### 2022-11-24 methodology.tex
added changes

Syn\footnote{\url{https://crates.io/crates/syn}} is a Rust crate built for the Rust procedural macro implementation. However, it includes a parser which is suitable for our purposes; we simply had to write a preprocessor to transform Syn AST output into Python dictionaries. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (\texttt{BCE}) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \texttt{BCE} elements can be found in Appendix~\ref{sec:appendix}.

@@

We propose a novel representation. Similar to the naive approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn. However, to simulate order-sensitiveness, we add two more steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs. Section~\ref{sec:weighting_scheme}, immediately below, illustrates this approach on an example..

@@

Our main requirement is that our datapoints summarize the key changes in ASTDiffs---a datapoint should foreground the most important change in a diff. For instance, the purple (right-most) path in Figure~\ref{fig:extraction} could be described as a change in a local variable declaration; a change in a function call; calling a method of one of the function arguments; or calling clone() on one of the function arguments. The last description is the most useful one, and we designed our weighting scheme to embed this value judgment.
We made two observations about changes. First, we saw that items (non terminals or BC-related elements) that occurs closer to AST leaves tend to be more semantically important. Second, and relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

@@

Furthermore, as we wanted to make sure that our tool captures patterns related to the Rust borrow checker, we manually increased the weights of the \texttt{BCE} items. The manual adjustment is why we characterize our weighting scheme as semi-automatically designed.

@@

Next, we collect a pair of revisions for each Rust file. The pair includes the state of the file before the commit and after the commit ($f_a, f_b$). After parsing each revision and computing the ASTDiff, we can compute the fixed sized datapoint $\mathit{DP}$. If the datapoint contains borrow-checker related keywords, we put it in the BC-related database $D_b$; otherwise, we put it in the general code changes database $D_g$. In both cases, we augment the datapoint with the commit hash, filename, and the scope in which the change happened. For BC-related code changes, we also store the detected BC-related keywords. Now our databases are ready for clustering and categorization.

@@

In this work, we were only interested in the first two groups and ignored clusters for refactoring changes. Also, we prefer reporting clusters that manifest more-specific patterns. An abstract pattern as opposed to a more-specific pattern can account for many changes in the programs. For instance, we deem the pattern `Changing a clone of a variable to a borrowing of it in a function argument' more specific than `Changing a statement in a function's body':

>>>>

Syn\footnote{\url{https://crates.io/crates/syn}} is a Rust crate built for the Rust procedural macro implementation. However, it includes a parser which is suitable for our purposes; we simply had to write a preprocessor to transform Syn AST output into Python dictionaries. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (\texttt{BCE}) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \texttt{BCE} elements can be found in our replication package.

@@

We propose a novel representation. Similar to the naive approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn (the columns are constant across all Rust programs). However, to simulate order-sensitiveness, we add two more steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs. Section~\ref{sec:weighting_scheme}, immediately below, illustrates this approach on an example..

@@

Our main requirement is that our datapoints summarize the key changes in ASTDiffs---a datapoint should foreground the most important change in a diff. For instance, the purple (right-most) path in Figure~\ref{fig:extraction} could be described as a change in a local variable declaration; a change in a function call; calling a method of one of the function arguments; or calling clone() on one of the function arguments. The last description, in our opinion, is the most useful one, and we designed our weighting scheme to embed this value judgment.
We made two observations about changes. First, we saw that items (non terminals or BC-related elements) that occurs closer to AST leaves tend to be more semantically important. Second, and relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

@@

Furthermore, as we wanted to make sure that our tool captures patterns related to the Rust borrow checker, we manually increased the weights of the \texttt{BCE} items. The manual adjustment is why we characterize our weighting scheme as semi-automatically designed. The final weights can also be found in our replication package.

@@

Next, we collect a pair of revisions for each Rust file. The pair includes the state of the file before the commit and after the commit ($f_a, f_b$). After parsing each revision and computing the ASTDiff, we can compute the fixed sized datapoint $\mathit{DP}$. If the datapoint contains borrow-checker related keywords, we put it in the BC-related database $D_b$; otherwise, we put it in the general code changes database $D_g$. In both cases, we augment the datapoint with the commit hash, filename, and the scope in which the change happened. For BC-related code changes, we also store the detected BC-related keywords. In summary, a datapoint is a tuple containing essence values for each element in $\mathtt{NT} \cup \mathtt{BCE}$ plus metadata about the change (e.g. commit hash, filename, etc.). Now our databases are ready for clustering and categorization.

@@

In this work, we were only interested in the first two groups and ignored clusters for refactoring changes. Also, we prefer reporting clusters that manifest more-specific patterns. An abstract pattern as opposed to a more-specific pattern can account for many changes in the programs. We use our judgment to prioritize patterns that we deem more specific. For instance, we deem the pattern `Changing a clone of a variable to a borrowing of it in a function argument' more specific than `Changing a statement in a function's body':

### 2022-11-24 related.tex
added changes

We would characterize our work as the Rust version of what was done in~\cite{hanam2016discovering} and~\cite{yang2022mining}. Like~\cite{hanam2016discovering}, we used DBSCAN to cluster fixes. However, our novel program embedding approach differs from what they did. In~\cite{hanam2016discovering}, they introduced Feature Properties table, which is a categorization for program elements. Embedding programs based on this table resulted in a order-sensitive embedding, however, it made the datapoints sparse. Moreover, like~\cite{yang2022mining}, we introduced patterns in two different categories. In our case there were general and language-specific patterns; for Rust, the language-specific patterns were related to the borrow checker (BC patterns). However, in oppose to our paper, in~\cite{yang2022mining}, they did a much more elaborate manual analysis, as in their data collection, they only collected general information about the single hunk changes (e.g. the number of variables, or arguments).

>>>>

We would characterize our work as the Rust version of what was done in~\cite{hanam2016discovering} and~\cite{yang2022mining}. Like~\cite{hanam2016discovering}, we used DBSCAN to cluster fixes. However, our novel program embedding approach differs from what they did. In~\cite{hanam2016discovering}, they introduced Feature Properties table, which is a categorization for program elements. Embedding programs based on this table resulted in a order-sensitive embedding, however, it made the datapoints sparse. Moreover, like~\cite{yang2022mining}, we introduced patterns in two different categories. In our case there were general and language-specific patterns; for Rust, the language-specific patterns were related to the borrow checker (BC patterns). However, as opposed to our paper, \cite{yang2022mining} needed to do a much more elaborate manual analysis, as they had only collected general information about the single hunk changes (e.g. the number of variables, or arguments).

### 2022-11-24 results.tex
added changes

We collected random samples 1000 times. In each iteration, we collected 50 sample datapoints from our general and borrow-checker databases (25 each). For each database, out of the 1000 sample sets, we picked the one that minimized the sampling error. The sampling error is defined as the absolute difference between the sample mean and the total mean. Also, these random samples are drawn from the clustered datapoints. The reason behind this decision is the same as why we collected a lot of noise points, that is, the majority of the commits involve changes in many program elements and the key changing elements are difficult to determine.

The human looks at the program output and assigns a rank to it

We carried out a manual experiment for evaluating our code embedding. In the experiment, the human analyzer looks at the provided circlepack figure and the change's source code and assigns a rank to it. This rank reflects the rank of the most important element in the circle pack figure according to the analyzer's view. Based on these ranks, then we calculate Top-$n$ Accuracy for $n=1$ up to $n=5$. Both authors acted as analyzer and carried out the experiment independently. We report the results as the fraction of the changes on which both authors agreed upon its embedding correctness. The final results are shown in Figure~\ref{fig:rq1}.

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, this modification also happens in three different ways. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information.

@@

Similar to other languages, such as C, a struct makes it possible for the users to define a custom type that is comprised of different types. The patterns that we describe in this section can also be seen in other languages that use the same concept. \\

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If a developer wants a variable to contain null, they would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that a variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field.

For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code to account for empty @media queries\footnote{\scriptsize \url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. As shown in the snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code, accordingly.

@@

Rust is a statically-typed language, and similar to the other statically-typed languages, some of the change patterns are associated with changes in types used in different program structures. Here, we introduce four bug fix patterns that are associated with types in Rust:

@@

\noindent\textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also, it can happen for refactoring purposes or performance enhancement.

For instance, in the Rust project starship/starship (a cross-shell prompt), in order to preserve the insertion order of \verb+substitutions+ field, the developer changes its type from \verb+HashMap+ to \verb+IndexMap+\footnote{\url{https://github.com/starship/starship/commit/4de9e43cff46c834bd340d24a02fc95d85310a33}}.
\\

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 8 borrow-checker related fix patterns, which we, similar to the previous table, describe in Table~\ref{table:bc}. Here we provide a detailed description for each of these patterns:

@@

\noindent\textbf{Description:} This is a change that in all instances has been detected by Clippy (commit messages indicated the use of Clippy for this fix). The change applies on an object prior to calling \verb+to_string()+ on that object. In this change, a dereference operator is added to that object. The reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods.

@@

\noindent\textbf{Description:} This change pattern removes the static keyword from a const variable declaration. This change can be applied using Clippy. If the presence of static is not required, it is better that to omit it, as keeping it might create complicated types in the program.

>>>>

We collected random samples 1000 times. In each iteration, 50 datapoints are sampled from our general and borrow-checker databases (25 each). For each database, out of the 1000 sample sets, we picked the one that minimized the sampling error. The sampling error is defined as the average of the absolute differences between the sample set mean and the total mean per column~\citep{degroot2012probability}. Also, these random samples are drawn from the clustered datapoints. The reason behind this decision is the same as why we collected a lot of noise points, that is, the majority of the commits involve changes in many program elements and the key changing elements are difficult to determine.

We carried out a manual experiment for evaluating our code embedding. In the experiment, the human analyzer looks at the change's source code and finds the most important element that appears in the circle pack figure, recording its rank. A value of 1 indicates that the human analyzer concurs with the algorithm, while 2 means that the human-identified most important element was ranked 2 by the algorithm. After collecting all the ranks, then we calculate Top-$n$ Accuracy for $n=1$ up to $n=5$. Both authors acted as analyzer and carried out the experiment independently. We report the results as the fraction of the changes on which both authors agreed upon its embedding correctness. The final results are shown in Figure~\ref{fig:rq1}.

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, Serde (A serialization/deserialization tool for Rust) calls a different function (\verb+is_empty+) to determine whether to skip serializing the \verb+attributes+ field.

@@

Like in other languages, such as C, a struct makes it possible for users to define a custom type that is comprised of different types. The patterns that we describe in this section can also be seen in other languages that use the same concept. \\

@@

\noindent\textbf{Description:} Rust is a null-safe language, meaning that object references cannot take on null values. If a developer wants a variable to contain the equivalent of null, they would have to use the \texttt{Option} type. \texttt{Option} is essentially an \texttt{enum} with two variants, \texttt{Some}, which indicates that a variable has a value, and \texttt{None}, which indicates the absence of any value. In this change pattern, the developer changes the type of a struct field to an \texttt{Option} of that type, hence enabling them to store no value in that field.

For instance, in the Rust project swc-project/swc (a fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code to account for empty @media queries\footnote{\scriptsize \url{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}. As shown in the snippet above, they changed type \texttt{MediaQueryList} to \verb+Option<MediaQueryList>+, and modified the other parts of the source code accordingly.

@@

Rust is a statically-typed language, and like in the other statically-typed languages, some change patterns are associated with changes in types used in different program structures. Here, we introduce four bug fix patterns that are associated with types in Rust:

@@

\noindent\textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also, it can happen for refactoring purposes or for performance enhancement.

For instance, in the Rust project starship/starship (a cross-shell prompt), to preserve the insertion order of the \verb+substitutions+ field, the developer changes its type from \verb+HashMap+ to \verb+IndexMap+\footnote{\url{https://github.com/starship/starship/commit/4de9e43cff46c834bd340d24a02fc95d85310a33}}.
\\

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 8 borrow-checker related fix patterns. Similar to the previous table, we describe BC-related patterns in Table~\ref{table:bc}. Now, we provide a detailed description of each of these patterns:

@@

\noindent\textbf{Description:} This is a change that in all instances (in our benchmarks) has been proposed by Clippy---commit messages indicated the use of Clippy for this fix. The change applies on an object prior to calling \verb+to_string()+ on that object. In this change, a dereference operator is added to that object. The reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods.

@@

\noindent\textbf{Description:} This change pattern removes the static keyword from a const variable declaration. This change can be applied using Clippy. A static lifetime of a variable indicates that the variable lives for the entire lifetime of the running program, and is not bound to a specific scope. If the presence of static is not required, it is better that to omit it, as keeping it might create complicated types in the program.
